---
title: "Test and benchmark of MsBackendParquet"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r style, echo = FALSE, results = "asis", message = FALSE}
BiocStyle::markdown()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BiocStyle)
```

# MsBackendParquet

In this notebook, we will test and benchmark the `MsBackendParquet` backend from
(Computional Metabolomics Research Group)[https://github.com/computational-metabolomics/MsBackendParquet]

```{r, eval = FALSE}
## BiocManager::install("MsBackendParquet")
pak::pak("computational-metabolomics/MsBackendParquet")
```

## Load libraries

```{r}
library(Spectra)
library(MsBackendParquet)
library(microbenchmark)
library(testthat)
```

## Test files

```{r}
file_1 <- MsDataHub::X20171016_POOL_POS_1_105.134.mzML()
files_2 <- c(MsDataHub::PestMix1_SWATH.mzML(), MsDataHub::PestMix1_DDA.mzML())
file_gz_1 <- MsDataHub::TMT_Erwinia_1uLSike_Top10HCD_isol2_45stepped_60min_01.20141210.mzML.gz()

files_4 <- c(
    MsDataHub::PestMix1_DDA.mzML(),
    MsDataHub::PestMix1_SWATH.mzML(),
    MsDataHub::X20171016_POOL_POS_1_105.134.mzML(),
    MsDataHub::X20171016_POOL_POS_3_105.134.mzML())
```

# Convert mzML to Parquet

First test is to convert the mzML file to Parquet format. With microbenchmark we
evaluate the efficiency of the process. I use the same parameter in the
vignette.

### Conversion `file_1` to Parquet format

```{r}
pq_file_1 <- "./parquet_file/file_1"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(file_1, path = pq_file_1,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `files_2` to Parquet format

The package generate a folder with inside a Parquet for each `mzML`.

```{r}
pq_files_2 <- "./parquet_file/files_2"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(files_2, path = pq_files_2,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `file_gz_1` to Parquet format

It handles the compressed files.

```{r}
pq_file_gz_1 <- "./parquet_file/file_gz_1"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(file_gz_1, path = pq_file_gz_1,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `files_4` to Parquet format

The package generate a folder with inside a Parquet for each `mzML`.

```{r}
pq_files_4 <- "./parquet_file/files_4"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(files_4, path = pq_files_4,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

# Test `MsBackendMzR` *vs* `MsBackendParquet`

In this section I will compere the standard `MsBackendMzR` backend for *raw* MS
data against the `MsBackendParquet` backend.

## `File_1`

Open file with `Spectra()`.

```{r}
microbenchmark(
    sps_mzr <- Spectra(file_1, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_file_1, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on `MsBackendParquet` is smaller
compared to the `MsBackendMzR`.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

We also evaluate extracting peaks data from selected spectra.

```{r}
set.seed(123)
idx <- sample(seq_along(sps_mzr), 13)
microbenchmark(
    a <- sps_mzr[idx],
    b <- sps_parq[idx],
    times = 10L, unit = "s")

microbenchmark(
    A <- peaksData(a),
    B <- peaksData(b),
    times = 10L, unit = "s"
)
expect_equal(A, B)
```

The subset operation is faster on the `MsBackendParquet` (since less data needs
to be subset). Extracting peaks data from the subset is however slower with `MsBackendParquet`.

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to `MsBackendMzR` for selected spectra variables.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(
        sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(
        sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)
```

On the other hand, if we read the full data with `spectraData()`, performance is
about the same (also for a data subset; not shown).

```{r}
microbenchmark(
    a <- spectraData(sps_mzr),
    b <- spectraData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(a[, intersect(colnames(a), colnames(b))],
             b[, intersect(colnames(a), colnames(b))])
```


Replace functions for peaks data (*m/z* or intensity) are not applied
immediately but cached in the `Spectra` object. Thus, these are independent of
the underlying backend. Replacement functions of spectra variables on the other
hand will be applied immediately, in the case of `MsBackendMzR` adding/replacing
the values in the internal `data.frame` with all spectra metadata, in the case
of the `MsBackendParquet` (presumably) an internal cache.

```{r}
rt <- rtime(sps_mzr)
microbenchmark(
    sps_mzr$rtime <- rt + 10,
    sps_parq$rtime <- rt + 10,
    times = 10L, unit = "s"
)

expect_equal(rtime(sps_mzr), rtime(sps_parq))
```

Replacing values in the `MsBackendParquet` are faster.

## `Files_2`

Open files with `Spectra()`. `files_2` contain 2 mzML.

```{r}
microbenchmark(
    sps_mzr <- Spectra(files_2, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_files_2, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
use more memory respect MzR.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Since the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## `File_gz_1`

Open file with `Spectra()`. `file_gz_1` contain a compressed mzML. In this case,
the Parquet method is way faster since the mzML is compressed.

```{r}
microbenchmark(
    sps_mzr <- Spectra(file_gz_1, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_file_gz_1, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
also in this case use more memory.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Sice the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## `Files_4`

Open files with `Spectra()`. `files_4` contain 4 mzML.

```{r}
microbenchmark(
    sps_mzr <- Spectra(files_4, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_files_4, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
use more memory respect MzR.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Since the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## Summary

# Comparing `MsBackendSql` *vs* `MsBackendParquet`

Both the `MsBackendSql` and `MsBackendParquet` store the MS data in their
respective own format, the former in a single SQL database, the latter in per
data-origin Parquet files.

We need to first create a SQL database for the `MsBackendSql`. For the tests we
will use the *largest* data set since both backends are developed for
large-scale data.

```{r}
library(MsBackendSql)

#' Create the SQLite database
library(RSQLite)
dbc <- dbConnect(SQLite(), "./files_4.sqlite")
createMsBackendSqlDatabase(dbc, files_4)
dbDisconnect(dbc)

sps_sql <- Spectra(dbname = "files_4.sqlite", source = MsBackendOfflineSql(),
                   drv = SQLite())
sps_parq <- Spectra(pq_files_4, source = MsBackendParquet())

print(object.size(sps_sql), units = "KB")
print(object.size(sps_parq), units = "KB")
```

Both backends have a very low memory footprint. Next we compare the time to get
the **full** spectra metadata data from the `Spectra` objects.

```{r}
microbenchmark(
    a <- spectraData(sps_sql),
    b <- spectraData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(a[, !colnames(a) == "dataStorage"],
             b[, !colnames(b) == "dataStorage"])
```

Extracting spectra data from `MsBackendParquet` is faster (maybe because there
is some caching involved?).

We evaluate also extraction of a single spectra variable.

```{r}
microbenchmark(
    a <- rtime(sps_sql),
    b <- rtime(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(a, b)
```

Also for a single spectrum variable `MsBackendParquet` is considerably faster.

We next compare extracting peaks data.

```{r}
microbenchmark(
    a <- peaksData(sps_sql),
    b <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(a, b)
```

Performance is very similar for both backends.

Next we compare the time it takes to subset the `Spectra` objects to random
spectra.

```{r}
set.seed(123)
idx <- sample(seq_along(sps_sql), 23)

microbenchmark(
    a <- sps_sql[idx],
    b <- sps_parq[idx],
    times = 10L, unit = "s"
)
```

Since both essentially just subset the spectra's primary keys their performance
is comparable.

Next we evaluate the time it takes to extract the **full** spectra data.

```{r}
microbenchmark(
    A <- spectraData(a),
    B <- spectraData(b),
    times = 10L, unit = "s"
)

expect_equal(
    A[, colnames(A) != "dataStorage"],
    B[, colnames(B) != "dataStorage"]
)
```

`MsBackendParquet` has performance advantages in extracting the full spectra
data.

We check also the performance for a single spectra variable:

```{r}
microbenchmark(
    A <- rtime(a),
    B <- rtime(b),
    times = 10L, unit = "s"
)

expect_equal(A, B)
```

Also for a single spectra variable `MsBackendParquet` is faster.

Evaluating extraction of the peaks data.

```{r}
microbenchmark(
    A <- peaksData(a),
    B <- peaksData(b),
    times = 10L, unit = "s"
)

expect_equal(A, B)
```

Here the `MsBackendSql` has performance advantages.

Checking if the order of spectra matters.

```{r}
idx <- sort(idx)
a <- sps_sql[idx]
b <- sps_parq[idx]

microbenchmark(
    A <- peaksData(a),
    B <- peaksData(b),
    times = 10L, unit = "s"
)
expect_equal(A, B)
```

The order of the spectra does not affect the performance of `MsBackendParquet`.


## Summary

- *spectra metadata*: `MsBackendParquet` is faster extracting the **full**
  spectra data and also single spectra variables. This performance advantage is
  present for the full data, but also for a subset of spectra in random order
  (same is true for an ordered subset of samples).
- Performance to extract the peaks data is about the same for both backends when
  the **full** data is used. For subsets of spectra (ordered or not)
  `MsBackendSql` is considerably faster (presumably because of the more
  efficient row extraction or the BLOB data type).

# Session information

```{r}
sessionInfo()
```
