---
title: "Test and benchmark of MsBackendParquet"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r style, echo = FALSE, results = "asis", message = FALSE}
BiocStyle::markdown()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BiocStyle)
```

# MsBackendParquet

In this notebook, we will test and benchmark the `MsBackendParquet` backend from
(Computional Metabolomics Research Group)[https://github.com/computational-metabolomics/MsBackendParquet]

```{r, eval = FALSE}
## BiocManager::install("MsBackendParquet")
pak::pak("computational-metabolomics/MsBackendParquet")
```

## Load libraries

```{r}
library(Spectra)
library(MsBackendParquet)
library(microbenchmark)
library(testthat)
```

## Test files

```{r}
file_1 <- MsDataHub::X20171016_POOL_POS_1_105.134.mzML()
files_2 <- c(MsDataHub::PestMix1_SWATH.mzML(), MsDataHub::PestMix1_DDA.mzML())
file_gz_1 <- MsDataHub::TMT_Erwinia_1uLSike_Top10HCD_isol2_45stepped_60min_01.20141210.mzML.gz()

files_4 <- c(
    MsDataHub::PestMix1_DDA.mzML(),
    MsDataHub::PestMix1_SWATH.mzML(),
    MsDataHub::X20171016_POOL_POS_1_105.134.mzML(),
    MsDataHub::X20171016_POOL_POS_3_105.134.mzML())
```

# Convert mzML to Parquet

First test is to convert the mzML file to Parquet format. With microbenchmark we
evaluate the efficiency of the process. I use the same parameter in the
vignette.

### Conversion `file_1` to Parquet format

```{r}
pq_file_1 <- "./parquet_file/file_1"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(file_1, path = pq_file_1,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `files_2` to Parquet format

The package generate a folder with inside a Parquet for each `mzML`.

```{r}
pq_files_2 <- "./parquet_file/files_2"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(files_2, path = pq_files_2,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `file_gz_1` to Parquet format

It handles the compressed files.

```{r}
pq_file_gz_1 <- "./parquet_file/file_gz_1"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(file_gz_1, path = pq_file_gz_1,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

### Conversion `files_4` to Parquet format

The package generate a folder with inside a Parquet for each `mzML`.

```{r}
pq_files_4 <- "./parquet_file/files_4"
microbenchmark({
    suppressMessages(be <- mzMLToParquet(files_4, path = pq_files_4,
                        partitioning = "dataOrigin", overwrite = TRUE))
}, times = 10L, unit = "s")
be
```

# Test MzR vs Parquet

In this section I will compere the standard method against MsBackendParquet

## `File_1`

Open file with `Spectra()`.

```{r}
microbenchmark(
    sps_mzr <- Spectra(file_1, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_file_1, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
continue to use less memory.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Sice the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## `Files_2`

Open files with `Spectra()`. `files_2` contain 2 mzML.

```{r}
microbenchmark(
    sps_mzr <- Spectra(files_2, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_files_2, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
use more memory respect MzR.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Since the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## `File_gz_1`

Open file with `Spectra()`. `file_gz_1` contain a compressed mzML. In this case,
the Parquet method is way faster since the mzML is compressed.

```{r}
microbenchmark(
    sps_mzr <- Spectra(file_gz_1, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_file_gz_1, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
also in this case use more memory.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Sice the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```

## `Files_4`

Open files with `Spectra()`. `files_4` contain 4 mzML.

```{r}
microbenchmark(
    sps_mzr <- Spectra(files_4, source = MsBackendMzR()),
    sps_parq <- Spectra(pq_files_4, source = MsBackendParquet()),
    times = 10L, unit = "s"
)
```

Comparing the (small) size, the object based on MsBackendParquet is smaller
respect the MzR one.

```{r}
print(object.size(sps_mzr), units = "KB")
print(object.size(sps_parq), units = "KB")
```

If we compare `spectraVariables()` between MzR and Parquet there is a
difference. In the parquet there is an additional variables.

```{r}
setdiff(spectraVariables(sps_parq), spectraVariables(sps_mzr))
```

Next I compare the extraction of the peaks using `peaksData()`. The parquet
method seems faster respect MzR. The function `expect_equal()` confirm that the
results are the same.

```{r}
microbenchmark(
    pks_mzr <- peaksData(sps_mzr),
    pks_parq <- peaksData(sps_parq),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr, pks_parq)
```

In extracting the data using `spectraData()` the parquet seems a bit slower
compared to MzR.

```{r}
microbenchmark(
    spsData_mzr <- spectraData(sps_mzr, columns = c("msLevel", "spectrumId", "dataStorage")),
    spsData_parq <- spectraData(sps_parq, columns = c("msLevel", "spectrumId", "dataStorage")),
    times = 10L, unit = "s"
)

expect_equal(spsData_mzr, spsData_parq)

```

On the other hand, if we read the full data with `spectraData()`, the Parquet
format is faster.
In this case, the `expect_equal()` check fail due to the additional column
detected before.

```{r}
microbenchmark(
    spsData_mzr_full <- spectraData(sps_mzr, columns = union(spectraVariables(sps_mzr),
                                        peaksVariables(sps_mzr))),
    spsData_parq_full <- spectraData(sps_parq, columns = union(spectraVariables(sps_parq),
                                        peaksVariables(sps_parq))),
    times = 10L, unit = "s"
)

## expect_equal(spsData_mzr_full, spsData_parq_full)

```

I filter the data, the filter function took ~same time, the parquet object
use more memory respect MzR.

```{r}
microbenchmark(
    sps_mzr_sub <- filterValues(sps_mzr, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    sps_parq_sub <- filterValues(sps_parq, spectraVariables = c("basePeakMZ", "rtime"),
                            values = c(123.089, 230), tolerance = c(0,5),
                            ppm = c(30, 0), match = "all"),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_sub), units = "KB")
print(object.size(sps_parq_sub), units = "KB")
```

Since the objects are smaller, also the difference in extracting the peaks is
smaller.

```{r}
microbenchmark(
    pks_mzr_sub <- peaksData(sps_mzr_sub),
    pks_parq_sub <- peaksData(sps_parq_sub),
    times = 10L, unit = "s"
)

expect_equal(pks_mzr_sub, pks_parq_sub)
```

The replace function is in favour of the Parquet file format. Took less time and
use less space.

```{r}
microbenchmark(
    sps_mzr_rep <- replaceIntensitiesBelow(sps_mzr, threshold = 10, value = 0),
    sps_parq_rep <- replaceIntensitiesBelow(sps_parq, threshold = 10, value = 0),
    times = 10L, unit = "s"
)

print(object.size(sps_mzr_rep), units = "KB")
print(object.size(sps_parq_rep), units = "KB")
```
